---
title: "Dam-building boom in the Mekong"
author: "Wei Jing Ang"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    theme: sandstone
    toc_float: true
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
bibliography: References.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup_packages, warning = FALSE, message = FALSE}
# Load libraries
library(tidyverse)
library(sf)
library(ggplot2)
library(patchwork)
library(ghibli)
```

#### Big Idea (to inform the public): The dam network in the Mekong is expected to expand by 60% over the next decade, doubling hydropower generation and positioning Laos as the next hotspot for dam development.

```{r load_data}
# Load data
# Mekong river basin
countries <- read_sf("Data/Mekong_basin_countries/Mekong_basin_countries.shp")
basin <- read_sf("Data/Mekong_basin_delta/Mekong_basin_delta.shp")

# Dams
dams <- read_sf("Data/Dams_1055/Dams_1055.shp")
dams_main <- read_sf("Data/Dams_mainstream/Dams_mainstream.shp")
dams1980 <- read_sf("Data/1980/1980.shp") # 1980
dams1990 <- read_sf("Data/1990/1990_combi.shp") # 1990
dams2000 <- read_sf("Data/2000/2000_combi.shp") # 2000
dams2010 <- read_sf("Data/2010/2010_combi.shp") # 2010
dams2020 <- read_sf("Data/2020/2020_combi.shp") # 2020

# River and lake
river1 <- read_sf("Data/River_1/FLOW1_STRA_3_China.shp") # upper course
river2 <- read_sf("Data/River_2/FLOW1_STRA_5.shp") # middle course
river3 <- read_sf("Data/River_3/FLOW3_CLASS2_TS.shp") # lower course
lake <- read_sf("Data/Lake/HydroLAKES_TS_finer.shp") # Tonle Sap Lake
```

The Mekong River, the once pulsing artery of Southeast Asia, is now clogged by a 
massive network of dams. In just the last ten years, over 200 dams have sprung up 
across the river basin, and the current dam network is projected to exceed a 
thousand within the next decade [@ang2023dams]. Among the planned projects are 
16 major dams that will double the length of the dam cascade straddling the 
mainstream Mekong. 

```{r dinos3, message = FALSE, warning = FALSE, out.width = "100%", fig.align = "center"}
# Base map of mekong basin
hbase <- ggplot() +
  geom_sf(data = basin, fill = "#CDE1E5", color = "transparent") + # lower basin
  geom_sf(data = rbind(river1, river2, river3), color = "#3981A4") + # river
  geom_sf(data = lake, fill = "#3981A4", color = "transparent") + # Tonle Sap Lake
  theme_void() +
  theme(legend.position = "inside",
        legend.position.inside = c(0.3,0.15)) # legend at bottom left

# With operating/ constructing dams
h1 <- hbase + 
  geom_sf(data = dams %>% 
            filter(Status == "Operational" | Status == "Under construction"),
          pch = 22, aes(fill = "#6A4E8E"), color = "white", size = 1.8, stroke = 0.3) + 
          # pch 22: square markers, color: border, size: marker size, stroke: border thickness
  scale_fill_manual(values = "#6A4E8E",
                    name = "",
                    label = "Operating\nConstructing")

# With planned dams
h2 <- hbase + 
  geom_sf(data = dams %>% 
            filter(Status == "Planned"),
          pch = 22, aes(fill = "#C66125"), color = "white", size = 1.8, stroke = 0.3) +
  scale_fill_manual(values = "#C66125",
                    name = "",
                    label = "Planned")

# Mainstream dams
h3 <- hbase + 
  geom_sf(data = dams_main %>% 
            filter(Status == "Operational" | Status == "Under construction"),
          pch = 22, fill = "#6A4E8E", color = "white", size = 1.8, stroke = 0.3) + 
  geom_sf(data = dams_main %>% 
            filter(Status == "Planned"),
          pch = 22, fill = "#C66125", color = "white", size = 1.8, stroke = 0.3) + 
  annotate("text", x = 99.9, y = 10.7, label = "Mainstream", size = 9/.pt) # mainstream label

h1 + h2 + h3 + plot_annotation(
  caption = "Source: Dams were compiled from 18 dam databases (Ang et al. 2023).\nThe blue area represents the Mekong River Basin, which includes the mainstream Mekong River and its tributaries."
) & 
  theme(plot.caption = element_text(hjust = 0))
```

# Hydropower generation will double

The Mekong is Southeast Asia's longest river, flowing through six developing 
countries: China, Myanmar, Laos, Thailand, Cambodia, and Vietnam [@hirsch2016shifting]. 
With these nations undergoing rapid economic and population growth, dams - as a 
relatively low-cost energy source in a region rich in water - have proliferated 
across the Mekong in response to rising energy demands [@hecht2019hydropower; @IRENA]. 

The total hydropower capacity, or the maximum power that can be generated by the 
entire dam network, has expanded exponentially from one gigawatt in the 1980s, 
to 35 gigawatts in the 2010s, and is projected to double to 70 gigawatts in the 2020s 
and beyond. 

China was the first to harness the Mekong's hydropower potential, as evidenced by 
its substantial expansion in hydropower capacity from the 2000s to the 2010s 
[@cronin2009mekong]. Having built 11 dams along the upper segment of the 
mainstream Mekong, China had become the focal point of discussions regarding the 
Mekong Basin, with many citing concerns about the impact of upstream dams on 
downstream countries [@brianeyler]. However, the pace of dam construction in China 
is slowing as development gains momentum downstream. 

```{r bar, message = FALSE, warning = FALSE, out.width = "100%", fig.align = "center"}
# Define function to sum hydropower per country per decade
bycountry <- function(hpyear,year){
  hpyear %>% 
  st_drop_geometry() %>%
  dplyr::select(Capacity_M, Country) %>%
  group_by(Country) %>%
  summarise(Capacity = sum(Capacity_M, na.rm = T)) %>%
  mutate(decade = year)
}

# Run function to sum hydropower
# Note: Myanmar only appears in 2010, so add in a row for Myanmar until 2000
country1980 <- bycountry(dams1980,1980) %>%
  rbind(.,data.frame(Country = "Myanmar",Capacity = 0, decade = 1980)) 
country1990 <- bycountry(dams1990,1990) %>%
  rbind(.,data.frame(Country = "Myanmar",Capacity = 0, decade = 1990))
country2000 <- bycountry(dams2000,2000) %>%
  rbind(.,data.frame(Country = "Myanmar",Capacity = 0, decade = 2000))
country2010 <- bycountry(dams2010,2010)
country2020 <- bycountry(dams2020,2020)
countryall <- rbind(country1980,country1990,country2000,country2010,country2020)

# Reorder the countries for a more organised bar graph
countryall$Country <- factor(countryall$Country, 
                             levels = c('Myanmar','Cambodia','Vietnam','Laos',
                                        'China','Thailand'),ordered = TRUE)

# Define palette
palette1 = ghibli_palette("MononokeMedium",direction=1)[c(2,5,7,4,3,6)]

# Bar graph of hydropower per country over time
h4 <- ggplot(countryall, aes(x = decade, y = Capacity, fill = Country)) +
  geom_col(width = 5) + # define width of bars
  scale_y_continuous(limits = c(NA,70000),
                     breaks=c(0,10000,20000,30000,40000,50000,60000,70000),
                     labels=c(0,"",20,"",40,"",60,"")) + # GW for cleaner graph
  scale_x_reverse(labels = c("1980s","1990s","2000s","2010s","Post-2020s")) + # ascending order
  scale_fill_manual(values = alpha(palette1,0.85)) + 
  xlab("") + ylab("Hydropower capacity (GW)") +
  theme(axis.ticks = element_blank(),
        axis.title = element_text(size=10),
        axis.text = element_text(size=10),
        panel.background = element_blank(),
        panel.grid.major.x = element_line("grey80", size = 0.1),
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank(),
        plot.background = element_blank(),
        legend.position = "none",
        aspect.ratio = 2/3) + 
  coord_flip() # horizontal bar graph
```

```{r country_map, message = FALSE, warning = FALSE, out.width = "100%", fig.align = "center", fig.height = 4}
# Reorder the countries from upstream to downstream
countries$NAME <- factor(countries$NAME,
                         levels = c('China','Myanmar','Laos','Thailand',
                                    'Cambodia','Vietnam'), ordered = TRUE)

# Define palette
palette2 = ghibli_palette("MononokeMedium",direction=1)[c(3,2,4,6,5,7)]

# Map of countries, with colours corresponding to the bar graph
h5 <- ggplot()+
  geom_sf(data = basin, color = "transparent", fill = alpha(palette2[6],0.85)) + # delta in Vietnam
  geom_sf(data = countries, aes(fill = NAME), color = "transparent") + 
  scale_fill_manual(values = alpha(palette2,0.85), name = "") + # no legend title
  theme_void() + 
  theme(
  legend.position = "inside",
  legend.position.inside = c(0.8,0.8), # legend at top right
  legend.key.height = unit(0.4, "cm"), # reduce legend key size
  legend.key.width = unit(0.4, "cm")
)

# Combine bar graph and map of countries
h4 + h5 + plot_annotation(
  caption = "Source: Ang et al. 2023. Post-2020s data includes dams under construction and all planned dams in the dam database."
) & 
  theme(plot.caption = element_text(hjust = 0))
```

# Laos as the next dam hotspot

With an ambitious goal of becoming the "Battery of Southeast Asia", Laos has the 
most planned dams and the highest expected growth in hydropower capacity over the 
next decade. Laos was the third-largest exporter of electricity globally in 2020, 
with 90% imported by Thailand, and its exports are projected to more than triple 
by 2030 [@chowdhury2020expected; @obsecon]. 

As shown in the sequence of maps below, Laos' hydropower capacity is expected to 
grow significantly in the 2020s and beyond. The volume of river flow increases 
downstream as more rainfall is captured, and Laos' location along the lower 
mainstream Mekong provides access to greater flow volumes that can be harnessed 
for hydropower [@biba2012china].

```{r dinos5, message = FALSE, warning = FALSE, out.width = "100%", fig.align = "center", fig.height = 3.5}
# Add geometry to countryall
countries <- countries %>% rename(Country = COUNTRY) # rename for joining
countryall_withgeom <- left_join(countryall,
                                 countries %>% 
                                   dplyr::select(geometry,Country),
                               by = "Country") %>% # left join geometries to country data
  st_as_sf(crs = st_crs(basin)) # turn into sf for mapping

# Decadal labels for the map
annotations <- data.frame(
  decade = c(1980, 1990, 2000, 2010, 2020),
  label = c("1980s", "1990s", "2000s", "2010s", "Post-2020s")
)

# Maps of hydropower capacity over the decades
ggplot()+
  geom_sf(data = basin, fill = "#D6D6FF", color = "transparent") + 
  geom_sf(data = countryall_withgeom, aes(fill = as.factor(ntile(Capacity, 5))), # quintiles
          color = "transparent") +
  facet_wrap(~decade, nrow = 1) + # 1 map for each decade
  scale_fill_manual(values = c("#D6D6FF","#B2B4F0","#6B7ED4","#4561BF","#003995"),
                    name = "GW",
                    labels = c("0","","","","33")) + 
  geom_sf(data = countries %>%
            filter(NAME == "Laos"), fill = "transparent", color = "#C66125", linewidth = 1) +
  geom_text(data = annotations, aes(x = 100, y = 8, label = label), size = 9/.pt) + # decadal labels
  theme_void() + 
  labs(caption = "Source: Ang et al. 2023. Laos is highlighted in red.\nPost-2020s data includes dams under construction and all planned dams in the dam database.") +
  theme(
  strip.text.x = element_blank(), # remove facet wrap labels
  legend.position = "inside",
  legend.position.inside = c(0.97,0.74), # legend at top right
  legend.key.height = unit(0.4, "cm"), # reduce legend key size
  legend.key.width = unit(0.4, "cm"),
  legend.title = element_text(size = 8),
  legend.text = element_text(size = 8),
  plot.caption = element_text(hjust = 0)
)
```

# Livelihoods and food security

While dams are viewed as an alternative energy source to fossil fuels, the surge 
in dam construction has become a threat to local livelihoods and regional food security. 
Dams fragment the river and obstruct migratory pathways for fish species, contributing 
to the concerning reality that one-fifth of the Mekong River's fish are now at risk of 
extinction [@wwf; @tang2021removing]. This is endangering the income of over 4 
million people in the world's largest inland fishery [@dugan2010fish]. 

Dams also block sediments from reaching the Mekong Delta, a vital agricultural 
region known as the "rice bowl of Vietnam", which accounts for 15% of global rice 
exports [@tri2012hydrology]. Starving the delta of its nutrient-carrying sediments 
is jeopardizing the agricultural industry [@grill2014development]. Laos' plans for 
7 new dams on the Mekong mainstream, along with Cambodia's plans for 2, could 
collectively prevent 96% of river sediments from reaching the delta 
[@kondolf2014dams]. Future dam management efforts should focus on Laos, the emerging
hotspot for hydropower development. 

# References
